---
title: "Data Visualization Practice & Notes"
author: "Colleen McCamy"
date: "2022-08-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)


library(tidyverse)
library(here)
library(janitor)
library(ggExtra)

# installing for ggplot

library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)
library(gapminder)
library(forcats)
library(sf)
library(lubridate)

```


### Reading in the lizard data

```{r}

lizards <- read_csv(here("data_tidy", "lizards.csv"))

```


## Making a heat map

```{r}

# Data Wrangling

lizard_counts <- lizards |> 
  mutate(date = mdy(date)) |> 
  count(year = lubridate::year(date), common_name) |> 
  drop_na()

  # doing both steps in one column
  # drop_na - drops rows with missing values
```


```{r}

ggplot(data = lizard_counts, aes(x = year, y = common_name)) +
  geom_tile(aes(fill = n), show.legend = FALSE) + 
  geom_text(aes(label = n), color = "white", size = 3) +
  scale_fill_gradientn(colours = c("navy", 'red', "orange", "yellow")
  )


# geom_text is combining a table and a plot with  adding the signs
# changing the color gradient with scale_fill_gradientn to be able to determine the number of colors to use

```


## Making a Beeswarm plot

```{r}

# data wrangling

whiptails <- lizards |> 
  filter(common_name == "western whiptail") |> 
  drop_na(total_length, weight)

# making the beeswarm plot
whiptail_bee <- ggplot(data = whiptails, aes(x = sex, y = weight)) +
  geom_beeswarm() +
  geom_boxplot(fill = NA)

```

# Making a Marginal Plot

```{r}


whiptails <- lizards |> 
  filter(common_name == "western whiptail") |> 
  drop_na(total_length, weight)

  
whiptail_plot <- ggplot(data = whiptails, aes(x = weight, y = total_length)) +
  geom_point(aes(color = sex)) +
  theme(legend.position = "bottom") + 
  theme_minimal()


ggMarginal(whiptail_plot, type = "boxplot", groupColour = TRUE)

# ggMarginal which is in the ggextra package - note that groupColour is with camel case and british 

```


## Patchwork plot

```{r}

((whiptail_bee + whiptail_plot) / whiptail_bee) & theme_dark()


# Patchwork results give you programatic results to have plot panels side by side - 

```


### GG Plot Using Data from Lizards

```{r}

ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point()

# can't switch to any graph - it must be compatible with the data that you have

```


Histogram of total lengths of all of the lizards in the data set

```{r}

plot_1_histogram <- ggplot(data = lizards, aes(x = total_length)) +
  geom_histogram(color= "#6E24D3", 
                 fill = "#BE99F0",
                 size = 0.5,
                 linetype = "dotted") +
  labs(x = "Total Length (mm)", y = "Count of Lizards") 

plot_1_histogram

# size refers to line width

# you don't want to change an aesthetic based on a constant in aes will map on to a variable value and bring up a legend, putting an extra legend that is mapped on to a variable but doesn't have meaning of what it is expected

```

Fill and color aesthetics with points

```{r}

ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(
    shape = 22,
    fill = "#6FF3BD",
    color = "#138560",
    size = 8,
    alpha = 0.4
  )

# alpha for transparence 0 is completely transparent 1 is opaque

```

Mapping Aesthetic based on variables

```{r}

ggplot(data = lizards, 
       aes(x = total_length, 
           y = weight)) + 
  geom_point(aes(color = common_name, 
                 size = total_length),
             alpha = 0.5) +
  theme_minimal()

```

Faceting plots


```{r}

ggplot(data = lizards, 
       aes(x = total_length, 
           y = weight)) + 
  geom_point(color = "#178893",
             alpha = 0.5) +
  facet_wrap(~common_name, scales = "free")

# faceting wraps the default is that the scale is the same for the x and y axis, however if you decide you don't care if the scales are not the same you can do as , scales = "free"

```

### Reordering things

```{r}

#finding the count of the lizards via group by and summarize

lizards_count <- lizards |> 
  group_by(common_name) |> 
  summarize(
    counts = n()
    )

# there is another way to do the same thing as above with a wrapper

lizards_count <- lizards |> 
  count(common_name)

# can do multiple variables at once as below

lizards_counts_ex <- lizards |> 
  count(common_name, site)

```

# Column graph that plots count or the value n - geom bar will do the counts of groups for you

```{r}

# almost always use geom_col to do a count plot

ggplot(data = lizards_count, 
       aes( y = fct_reorder(common_name, n),
            x = n)) +
  geom_col(fill = "#178893") + 
  labs(x = "Count",
       y = "Lizard Common Name")

# default in ggplot with a character variable is to be plotted in alphabetical order 

# for a character variable cannot specify the order things appear in

# using a new function - allow you to reorder things, don't want to manually code in the orders

# using function `fct_reorder()` 

# taking common_name to make it a factor and need to tell it the variable to determine the order

```

### Creating scatterplots faceted by species for total length vs. weight. The color and shape of points should differ based on species(common name.) Change the theme and Make each faceted panel have flexible axis limits

```{r}

ggplot(data = lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point(aes(color = common_name), 
             alpha = 0.5,
             shape = 22) + 
  facet_wrap(~common_name, scales = "free") +
  theme_minimal() +
  labs( x = "Total Length (mm)", y = "Weight (grams)", 
        title = "Jornada Basin Lizard Sizes",
        subtitle = "All lizards follow the standard length-weight relationship, weith Western Whiptails being largest",
        caption = "Data : Jordana Basin LTER, CCBY")

# can manually create a vector for shapes shape = c(22, 14, 5, 13, 17)

# for facet wrap free can do scales = "free.x" or scales = "free.y" to only have it free in one direction

```

## Using Scales to Add Customizations

```{r}

ggplot(data = lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_gradientn(colors = c("magenta", "#FD52C9", "#20B95D", "#144591", "dodgerblue"))


# scale_color_gradient = two colors
# scale_color_gradient = colors = c()

```




```{r}

ggplot(data = lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_steps(low = "#EEC435", high = "#35EE7B")

#scale_color_steps = based on steps of breakpoints ^ automatic ones


ggplot(data = lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point(aes(color = total_length)) +
  scale_color_steps2(low = "#EEC435", 
                     mid = "#4CAD2C",
                     high = "#360EBF",
                     midpoint = 150,
                     breaks = c(50, 75, 150, 180, 220, 280))

#scale_color_steps2 = tell what the high, low and mid values are for a diverting color scheme - creating a breakpoints at the different breaks


ggplot(data = lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point(aes(color = total_length)) +
  scale_color_stepsn(colors = c("orange", "red", "purple"),
                     breaks = seq(from = 0, to = 300, by = 60))

# instead of divering color schemes create sequences for break points
# what if this changes and you hard code these break points could say sequence and maximum value and divide by five and sequence along that for the iteration - do this outside of ggplot to then call it back in gg plot

# each addition is adding a new layer on to a ggplot so it looks a the most recent line of code for instance the scale is in reference to the point geom and if there was a line geom we would need to have another scale after the line plot

# should have a graph via a function - create a theme in gg plot and 

```

## Creating a function to use ggplot

```{r}

# you can create a theme in ggplot and make a custom theme and then include in a function and apply that function directly to a graph


```


## Palletteer & Factor Reorder `fct_reorder()`

```{r}

# palletts are just a vector of custom colors and items
# pallette collects all of the different pallettes so you can use them


# could make sense to have the version of the data stored in that order
lizards_median <- lizards |> 
  mutate(common_name = fct_reorder(common_name, 
                                   total_length, 
                                   .fun = median))

# this is creating a new column overwriting an exisiting column, recast column name as a factor and then to put those in order based on the median total length based on the median factor


ggplot(data = lizards, aes(x = total_length, y = common_name)) +
  geom_boxplot(aes(fill = common_name)) +
  theme_minimal()


ggplot(data = lizards_median, aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) +
  scale_fill_paletteer_d(palette = "LaCroixColoR::Pamplemousse") +
  theme_minimal()

# reorder from high to low and change the color scheme
# for the palette aggregate package of color palettes it builds into it- you need to identify the package and then package view to see more do View(palette_d_names)

```

## Making an awesomely awful graph

```{r}

ggplot(data = lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point(aes(color = weight)) +
  theme(panel.grid.major.x = element_line(color = "red"),
        panel.grid.minor.y = element_blank(), 
        axis.title = element_text(color = "purple", size = 10),
        axis.text.y = element_text(color = "green"),
        text = element_text(size = 18),
        panel.background = element_rect(color = "purple", 
                                        fill = "yellow")
        )


# theme the highest element of resolution will be the one that will be used - for instance if you edit the font size for all text and have a title for x axis

# element_rectangle for the graph

```

## Adding custom annotations and adding a line

```{r}

ggplot(data = lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point(aes(color = weight)) +
  theme_minimal() +
  annotate("text", x = 300, y = 50, label = "Wooooo", color = "purple") +
  geom_vline(xintercept = 250, 
             linetyple = "dashed", 
             color = "orange", 
             size = 4)
        )

# annotate ( say what you want, then where you want and then what to say and the color)

# customizing things are often in themes

# geom_vline - to add a line in the graph say where you want it, line type, color, size, etc

```


## Making repulusive labels with `ggrepel` and gapminder


```{r}

wws_lizards <- lizards |> 
  filter(common_name == "western whiptail", 
         site == "sand")

ggplot(data = wws_lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point() +
  geom_text_repel(aes(label = toe_num))

# you can use geom_text to add labels, default for geom text to center everything horizontally and vertically over the coordinate for that valye, which doesn't work well for having the point and the label with geom_text()

# geom_text_repel() this will update the formatting of the geom_text and will make the default to have automatic repulsion away from the point so that you can still read the values

```

### using `gghighlight()` and `geom_text_repel()`

```{r}

# doing a filter and then piping straight into gg plot

gapminder |> 
  filter(year == 2022 & continent == "Europe") |> 
  ggplot(aes(x = gdpPercap, y = lifeExp)) + 
  geom_point() +
  geom_text_repel(aes(label = country), size = 3)

# using h

ggplot(data = wws_lizards, aes(x = total_length, 
                           y = weight)) +
  geom_point() +
  gghighlight(toe_num == 250, label_key = toe_num)

# gghighlight first is logical askings is this true, then if it is truen then add a label to it

```

## Making a map using spatial data

```{r}

jornada_veg <- read_sf(here::here("data_raw", "spatial_vegetation", "doc.kml"))

# even with different types of data you can use ggplot in similar ways - don't need to specify x and y since it will recognize the spatial data

ggplot() +
  geom_sf(data = jornada_veg,
          aes(fill = Name),
          color = NA) +
  theme_minimal() +
  scale_fill_paletteer_d(palette = "ggthemes::manyeys")
# can't use read csv cause it is not a csv file read_sf -
# sf = simple figures with all of you attributes and the spatial information stays stuck to the data
```


